---
title: "General introduction"
author: "Anton Klåvus, Vilhelm Suksi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{General introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
biblio-style: apalike
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

# Motivation

From the perspective of metabolites as the continuation of the central dogma of biology, metabolomics provides the closest link to many phenotypes of interest. This makes metabolomics research promising in teasing apart the complexities of living systems. However, the high-dimensional data resulting from untargeted LC-MS metabolomics experiments requires sophisticated data science methodologies to achieve quality, reproducible results. 

The ```notame``` package provides functionality for untargeted LC-MS metabolomics research as specified in the associated publication in the "Metabolomics Data Processing and Data Analysis—Current Best Practices" special issue of the Metabolites journal [@notame]. This includes data pretreatment, quality control, uni- and multivariate analysis as well as common visualizations. Raw data preprocessing and functionality related to biological context, such as pathway analysis, is not included.

Bioconductor's emphasis on reproducibility and the streamlined ```notame``` experience makes for a solid introduction to quality, reproducible metabolomics research.

# Installation

To install ```notame```, install BiocManager first, if it is not installed. Afterwards use the install function from BiocManager.

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("notame")
library(notame)
```

# How it works

MetaboSet objects are the primary data structure of this package. MetaboSet is built upon the ExpressionSet class from the Biobase package by Bioconductor, extended to make for a streamlined experience. Separate MetaboSet objects are created for each analytical mode from a peak list file using ```read_from_excel``` and ```construct_metabosets``` to accommodate typical output from peak-picking software such as Agilent's MassHunter or MS-DIAL. This includes checks and manipulation of the metadata to prepare the object for the analysis workflow. 

In addition to basic ExpressionSet functionality, the following slots and functions of the same names extend the MetaboSet class: ```group_col```, ```time_col``` and ```subject_col```. They are used for accessing and setting names of columns with corresponding functionality, and are used as default arguments in many functions.

Data pretreatment and quality control encompasses many functions which return modified MetaboSet objects. Central functions returning modified MetaboSet objects include (in roughly chronological order):

- ```mark_nas``` for marking missing values as ```NA``` (peak picking software tend to report them as 0 or 1) 
- ```flag_detection``` for flagging features with abundant missing values (NAs), with different thresholds for biological samples and QC samples [@qcguidelines]
- ```flag_contaminants``` for flagging features detected in blank samples
- ```correct_drift``` for QC-based drift correction [@driftcorrection]
- ```assess_quality``` for feature quality metrics, that is RSD,  D-ratio and non-parametric equivalents [@qcguidelines]
- ```flag_quality``` for flagging features according to quality metrics
- ```cluster_features``` for clustering and aggregating redundant features originating from the same metabolite
- ```merge_metabosets``` for merging metabosets spanning multiple modes together
- ```ruvs_qc``` for batch correction [@ruvsnormalization]
- ```impute_simple``` and ```impute_rf``` for imputation
- ```pqn_normalize``` for Probabilistic Quotient Normalization [@pqn]
- ```scale``` for centering and/or scaling to unit variance
- ```drop_qcs``` for removing QC samples before statistical analysis
- ```drop_flagged``` for removing flagged features before statistical analysis
- ```join_fData``` for including results of statistical analyses in the object

Univariate statistics functions return a ```data.frame```, to be manually filtered before inclusion into the feature data of the instance. Multivariate statistics functions return various data structures. 

Visualization functionality can be broadly split into QC-visualizations, feature-wise visualizations and comprehensive results visualizations:
 
- QC-visualizations, which also serve as exploratory data analysis, are saved to file by default but can be returned. The ```visualizations``` wrapper can be used for saving QC-visualizations at different stages of data pretreatment.
- feature-wise visualizations are saved to file by default but can be returned as a list
- comprehensive results visualizations are returned 

For keeping track of the analysis, ```notame``` offers a logging system operated using ```init_log```, ```log_text``` and ```finish_log```. Other useful utilities include ```combined_data``` for representing the instance in a ```data.frame``` suitable for plotting, ```MakeSummarizedExperimentFromMetaboSet``` for conversion to capitalize on the interoperability of Bioconductor and ```write_to_excel``` to save the data in an instance to an Excel file.

Parallellization is provided by the ```BiocParallel``` package. ```BiocParallel``` defaults to a parallel backend which can be changed using ```BiocParallel::register```. For a platform-agnostic choice, do:

```{r, eval = FALSE} 
BiocParallel::register(BiocParallel::DoparParam())
```

```notame``` also keeps track of all the external packages used, offering you references for each. To see and log a list of references, use citations()

To capitalize on the reproducibility of Bioconductor and programmatic data analysis, it can make sense to code the analysis as a computational document (e.g. Quarto). Please see the other vignette for a project example. 

# Authors & Acknowledgements

The first version of ```notame``` was written by Anton Klåvus for his master's thesis in Bioinformatics at Aalto university (published under former name Anton Mattsson), while working for University of Eastern Finland and Afekta Technologies. The package is inspired by analysis scripts written by Jussi Paananen and Oskari Timonen. The algorithm for clustering molecular features originating from the same compound is based on MATLAB code written by David Broadhurst, Professor of Data Science & Biostatistics in the School of Science, and director of the Centre for Integrative Metabolomics & Computational Biology at the Edith Covan University.

If you find any bugs or other things to fix, please submit an issue on GitHub! All contributions to the package are always welcome!

# Session information

```{r, echo = FALSE}
sessionInfo()
```

# References